Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-04-12 ms/spaceness.tex
gwas text and fig revision

\subsection{Association Studies}
To assess the degree to which spatial structure confounds GWAS we simulated four types of nongenetic phenotype variation for 1000 randomly sampled individuals in each spatial SLiM simulation and conducted a linear-regression GWAS with PC covariates in PLINK \citep{PURCELL2007}. SNPs with a minor allele frequency less than 0.5\% were excluded from this analysis. Phenotype values were meant to roughly reflect the distribution of height across Europe \ak{how so? need more details here. Was it matched to mean and variance?}, which has recently been found to be confounded with population structure in large scale GWAS \citep{Berg2018,Sohail2018}. Conceptually our approach is similar to that taken in \citep{Mathieson2012}, though here we model fully continuous spatial variation and focus on genome-wide effects rather than low-frequency alleles specifically.

@@

\subsection{GWAS}
For simulations with neighborhood size under 100 spatial variation in the environment causes uncorrected GWAS to infer significant associations with purely environmental phenotypes at most genome-wide SNPs (Figure \ref{fig:gwas}). This effect is particularly strong for clinal and corner environments, for which the lowest dispersal levels cause over 95\% of segregating sites in the sample to return significant associations. Patchy environmental distributions, which result in fewer extreme phenotype values in the sample (Figure \ref{fig:gwas}A), cause fewer false-positives overall but still produce up to 1000 spurious associations. Notably less than 1\% of simulations with nonspatial environments returned any significant associations (Figure \ref{fig:gwas}C), demonstrating that this effect is caused specifically by the interaction of population structure and spatial variation in the environment rather than by population structure itself.

When PCA positions are included as covariates the vast majority of SNPs no longer pass a 5\% FDR significance threshold, but hundreds of genome-wide significant hits (out of an average of 132,000 SNPs) were observed even in relatively high dispersal simulations for the corner and clinal distributions (Figure \ref{fig:gwas}C). In most cases the $p$ values for these associations were significant after FDR correction but would not pass the threshold for significance under the more conservative Bonferroni correction, suggesting that the most strongly associated variants from many GWAS of mono- or oligogenic traits are robust to this variety of stratification bias. Though patchy environmental distributions produce fewer false positive associations overall, PC correction appears less effective as a correction for in this case -- dropping the count of significant sites by roughly half rather than $>95\%$ as in clinal or corner distributions.

Clinal environments cause an interesting pattern in the count of significant sites after PC correction: at low neighborhood sizes the correction appears to work well and removes nearly all significant associations, but at neighborhood sizes above $\approx250$ the number of significant SNPs increases to between 10 and 400 (Figure \ref{fig:gwas}. This appears to reflect a loss of descriptive power in the PCA -- as neighborhood size increases, the total proportion of variance explained by the first 10 PC axes declines from roughly 0.1 to 0.04 (Figure \ref{gwas}B). Essentially, PCA seems unable to effectively summarize the weak population structure present in large-neighborhood simulations, but these populations continue to have enough spatial structure to create significant correlations between genotypes and the environment. A similar process can also be seen in the corner phenotype distribution, in which the count of significant SNPs initially declines as neighborhood size increases and then increases at approximately the point at which the PCA reaches its nadir.

In addition to examining the count of significant SNPs we also explored the degree of genome-wide inflation of test statistics by plotting quantile-quantile plots for all simulations and phenotype distributions (Figure \ref{fig:gwas}D). For clinal environments $-log_{10}(p)$ values are most inflated when neighborhood sizes are large, consistent with the pattern observed in the count of significant associationsa after PC regression. In contrast corner and patchy environments cause the greatest inflation in $-log_{10}(p)$ at neighborhood sizes $<100$, which likely reflects the inability of PCA to account for fine-scale structure caused by very limited dispersal. We also observed that PC regression appears to cause some degree of overcorrection for all phenotype distributions, visible in Figure \ref{fig:gwas}D as points falling below the 1:1 line. This suggests that PC regression is removing some true signal of association in our simulations. This result is consistent with a recent empirical study of heritability in human height and body mass index, which found that increasing the number of PC axes used as covariates caused the total proportion of variance explained by SNPs to decline from $\approx0.8$ to $\approx0.75$ \citep{Wainschtein2019}. Indeed SNPs with minor alleles frequencies of 0.001 - 0.01, which are expected to reflect fine scale population structure \citep{Mathieson2012,Novembre2009}, are estimated to explain \textit{negative} proportions of the total phenotypic variance in \citep{Wainschtein2019}.

\section{Discussion}

Patterns of genetic variation are influenced by the fact that organisms are more likely, on average, to reproduce with others of their species that are geographically proximate, in particular when dispersal distances are short. While historically some analytical progress has been made in describing the influence of continuous space on population genetics (e.g. \cite{Wright1943,Rousset1997}), the theoretical work is challenging and often untenable for realistic biological models. Instead we here use efficient forward time population genetic simulations to describe the myriad influence of space on genetic variation. In particular we examine three axes of variation across four different sampling strategies-- 1) population genetic summary statistics, 2) inference of population size history, and 3) the consequences on genome-wide association studies (GWAS). We are in particular interested in asking how our empirical inferences from data might be affected by spatial processes. As we show below the answers seems to be- often space matters, both because of how populations are sampled with respect to space, and because of the inherent dispersal properties of those populations.

\subsection{Effects of Dispersal}
Recap the general effects of summary stats-- skew in SFS, inflation of pi for low dispersal. What did we know about this before i.e from literature?

@@

@@

\afterpage{\clearpage}
\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/stairwayplot_facet_rollmean.pdf}
\caption{Inferred demographic histories for spatial SLiM simulations from Stairwayplot, by sampling scheme and neighborhood size (NS) range. The thick line is a rolling mean and thin lines are individual model fits. Dashed horizontal lines are the average $N_{e}$ across random-mating SLiM models estimated from $\pi$.}
\label{fig:demography}
\end{figure*}

\afterpage{\clearpage}
\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/gwas_summary_nsig_qqplots_loglog.pdf}
\caption{Impacts of spatially correlated phenotypes and isolation by distance on linear regression GWAS.  A: example phenotype distributions and sampling maps, with colors scaled to phenotype values. B: proportion of total variance explained by the first 10 PC axes, by neighborhood size. C: numbers of significant SNPs after FDR correction for linear-model GWAS conducted with (top) or without (bottom) PC covariates. D: Quantile-quantile plots of observed $p$ values relative to those expected from a uniform distribution. The dotted lines show the 95\% confidence region assuming binomial sampling and an alpha level of 0.05. Points above the 1:1 line reflect deflated $p$ values.}
\label{fig:gwas}
\end{figure*}

\afterpage{\clearpage}
% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Thu Apr 11 13:11:59 2019
\begin{table}[ht]
\centering
\caption{\bf Neighborhood size estimates from empirical studies.}
\begin{tabular}{rllrrll}
  \hline
 Species & Description & \makecell[l]{Neighborhood\\Size} & Method & Citation \\ 
  \hline
  \makecell[l]{\textit{Ipomopsis}\\\textit{aggregata}} & flowering plant & 12.60-37.80 & Genetic & Campbell et al. 1992\\ 
  \makecell[l]{\textit{Linanthus}\\\textit{parryae}} & flowering plant & 14-27 & Genetic & Wright 1943\\ 
  \makecell[l]{\textit{Borrichia}\\\textit{frutescens}} & salt marsh plant & 20-30 & Genetic+Survey & Antlfinger 1982 \\ 
  \makecell[l]{\textit{Oreamnos}\\\textit{americanus}} & mountain goat & 36-100 & Genetic & Shirk and Cushman, 2014 \\ 
  \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{Gainj- and Kalam- \\speaking people,\\Papua New Guinea} & 40-213 & Genetic & Rousset et al. 1997\\ 
  \makecell[1]{\textit{Formica sp.}} & colonial ants & 50-100 & Genetic & Pamilo 1982 \\ 
  \makecell[l]{\textit{Astrocaryum}\\\textit{mexicanum}} & palm tree & 102-895 & Genetic+survey & Eguiarte et al. 1993 \\ 
  \makecell[l]{\textit{Spermophilus}\\\textit{mollis}} & ground squirrel & 204-480 & Genetic+Survey & Antolin et al. 2001\\
  \makecell[l]{\textit{Sceloperus}\\\textit{olivaceus}} & lizard & 225-270 & Survey & Kerster 1964 \\ 
  \makecell[l]{\textit{Dieffenbachia}\\\textit{longispatha}} & \makecell[l]{beetle-pollinated\\colonial herb} & 227-611 & Survey & Young 1988 \\ 
  \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{Gainj- and Kalam- \\speaking people,\\Papua New Guinea} & 410 & Survey & Rousset et al. 1997 \\ 
  \makecell[l]{\textit{Quercus}\\\textit{laevis}} & Oak tree & $\geq440$ & Genetic & Berg and Hamrick, 1994 \\ 
  \makecell[l]{\textit{Drosophila}\\\textit{pseudoobscura}} & fruit fly & 500-1000 & Genetic & Wright et al. 1941 \\ 
  \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{NE Europe} & 1342-5425 & Genetic & Ringbauer et al. 2018 \\ 
  \makecell[l]{\textit{Bebicium}\\\textit{vittatum}} & intertidal snail & 240000-240000 & Survey & Rousset et al. 1997\\ 
  \makecell[l]{\textit{Bebicium}\\\textit{vittatum}} & intertidal snail & 360000-360000 & Genetic & Rousset et al. 1997\\ 
   \hline
\end{tabular}
\label{table:NStable}
\end{table}

>>>>

\subsection{Association Studies}
To assess the degree to which spatial structure confounds GWAS we simulated four types of nongenetic phenotype variation for 1000 randomly sampled individuals in each spatial SLiM simulation and conducted a linear regression GWAS with PC covariates in PLINK \citep{PURCELL2007}. SNPs with a minor allele frequency less than 0.5\% were excluded from this analysis. Phenotype values were meant to roughly reflect the distribution of height across Europe \ak{how so? need more details here. Was it matched to mean and variance?}, which has recently been found to be confounded with population structure in large scale GWAS \citep{Berg2018,Sohail2018}. Conceptually our approach is similar to that taken in \citep{Mathieson2012}, though here we model fully continuous spatial variation and focus on genome-wide effects rather than low-frequency alleles specifically.

@@

\subsection{GWAS}
To ask what confounding effects spatial genetic variation might have on genome-wide association studies we performed GWAS on our simulations using phenotypes that were determined solely by the environment. In general we found that for simulations with limited dispersal, i.e. neighborhood size $< 100$, spatial variation in the environment causes GWAS to infer significant associations with purely environmental phenotypes at over 25\% of sampled SNPs if no correction for genetic relatedness among samples is performed (Figure \ref{fig:gwas}). This effect is particularly strong for clinal and corner environments, for which the lowest dispersal levels cause over 60\% of SNPs in the sample to return significant associations. Patchy environmental distributions, which result in fewer extreme phenotype values in the sample (Figure \ref{fig:gwas}A), cause fewer false-positives overall but still produce spurious associations at roughly 10\% of sites at the lowest neighborhood sizes. Notably no simulations with nonspatial environments returned more than one significant association (Figure \ref{fig:gwas}C), demonstrating that this effect is caused specifically by the interaction of population structure and spatial variation in the environment rather than by population structure itself.

When PCA positions are included as covariates for GWAS the vast majority of SNPs no longer pass a 5\% FDR significance threshold, but up to 1.5\% of SNPs are significantly associated at low dispersal distances under corner and patchy environmental distributions (Figure \ref{fig:gwas}C). At neighborhood sizes $> 500$ up to 0.31\% of SNPs were significant for corner and clinal environments. Given an average of 132,000 SNPs across simulations after MAF filtering, this translates to up to 382 false-positive associations. In most cases the $p$ values for these associations were significant after FDR correction but would not pass the threshold for significance under the more conservative Bonferroni correction, suggesting that the most strongly associated variants from many GWAS of mono- or oligogenic traits are robust to this variety of stratification bias.

Clinal environments cause an interesting pattern in the count of significant sites after PC correction: at low neighborhood sizes the correction removes nearly all significant associations, but at neighborhood sizes above $\approx250$ the proportion of significant SNPs increases to up to 0.4\% (Figure \ref{fig:gwas}. This appears to reflect a loss of descriptive power in the PCA -- as neighborhood size increases, the total proportion of variance explained by the first 10 PC axes declines from roughly 0.1 to 0.04 (Figure \ref{fig:gwas}B). Essentially, PCA seems unable to effectively summarize the weak population structure present in large-neighborhood simulations, but these populations continue to have enough spatial structure to create significant correlations between genotypes and the environment. A similar process can also be seen in the corner phenotype distribution, in which the count of significant SNPs initially declines as neighborhood size increases and then increases at approximately the point at which the proportion of variance explained by PCA approaches its minimum.

In Figure \ref{fig:gwas}D we present quantile-quantile plots that show the degree of genome-wide inflation of test statistics in PC-corrected GWAS across all simulations and environmental distributions. For clinal environments $-log_{10}(p)$ values are most inflated when neighborhood sizes are large, consistent with the pattern observed in the count of significant associations after PC regression. In contrast corner and patchy environments cause the greatest inflation in $-log_{10}(p)$ at neighborhood sizes $<100$, which likely reflects the inability of PCA to account for fine-scale structure caused by very limited dispersal. We also observed that PC regression appears to cause some degree of overcorrection for all phenotype distributions, visible in Figure \ref{fig:gwas}D as points falling below the 1:1 line. This suggests that PC regression is removing some true signal of association in our simulations.

\section{Discussion}
Patterns of genetic variation are influenced by the fact that organisms are more likely, on average, to reproduce with others of their species that are geographically proximate, in particular when dispersal distances are short. While historically some analytical progress has been made in describing the influence of continuous space on population genetics (e.g. \cite{Wright1943,Rousset1997,Ringbauer2017,Barton2010}), the theoretical work is challenging and often untenable for realistic biological models. Instead we here use efficient forward time population genetic simulations to describe the myriad influence of space on genetic variation. In particular we examine three axes of variation across four different sampling strategies-- 1) population genetic summary statistics, 2) inference of population size history, and 3) the consequences on genome-wide association studies (GWAS). We are in particular interested in asking how our empirical inferences from data might be affected by spatial processes. As we show below the answers seems to be- often space matters, both because of how populations are sampled with respect to space, and because of the inherent dispersal properties of those populations.

\subsection{Effects of Dispersal}
Limited dispersal inflates estimates of population size from genetic variation,

Recap the general effects of summary stats-- skew in SFS, inflation of pi for low dispersal. What did we know about this before i.e from literature?

@@

This result is consistent with a recent empirical study of heritability in human height and body mass index, which found that increasing the number of PC axes used as covariates caused the total proportion of variance explained by SNPs to decline from $\approx0.8$ to $\approx0.75$ \citep{Wainschtein2019}. Indeed SNPs with minor alleles frequencies of 0.001 - 0.01, which are expected to reflect fine scale population structure \citep{Mathieson2012,Novembre2009}, are estimated to explain \textit{negative} proportions of the total phenotypic variance in \citep{Wainschtein2019}.

@@

\afterpage{\clearpage}
\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/stairwayplot_facet_rollmean.pdf}
\caption{Inferred demographic histories for spatial SLiM simulations from Stairwayplot, by sampling scheme and neighborhood size (NS) range. The thick line is a rolling mean and thin lines are individual model fits. Dashed horizontal lines are the average $N_{e}$ across random-mating SLiM models estimated from $\theta_{\pi}$.}
\label{fig:demography}
\end{figure*}

\afterpage{\clearpage}
\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/gwas_summary.pdf}
\caption{Impacts of spatially correlated phenotypes and isolation by distance on linear regression GWAS.  A: example environmental distributions and sampling maps, with colors scaled to phenotype values. B: proportion of total variance explained by the first 10 PC axes, by neighborhood size. C: Proportion of significant SNPs after FDR correction for linear-model GWAS conducted with (top) or without (bottom) PC covariates, by neighborhood size. D: Quantile-quantile plots of observed $p$ values relative to those expected from a uniform distribution. The dotted lines show the 95\% confidence region assuming binomial sampling and an alpha level of 0.05. Points above the 1:1 line reflect deflated $p$ values.}
\label{fig:gwas}
\end{figure*}

\afterpage{\clearpage}
% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Thu Apr 11 13:11:59 2019
\begin{table}[ht]
\centering
\caption{\bf Neighborhood size estimates from empirical studies. Survey methods indicate field studies to measure dispersal or population density. Genetic methods include both sequence-based analyses and enzyme activity assays or crosses used to estimate allele frequencies.}
\begin{tabular}{rllrrll}
  \hline
 Species & Description & \makecell[l]{Neighborhood\\Size} & Method & Citation \\ 
  \hline
  \makecell[l]{\textit{Ipomopsis}\\\textit{aggregata}} & flowering plant & 12.60 - 37.80 & Genetic & \citep{Campbell1992}\\
  \makecell[l]{\textit{Linanthus}\\\textit{parryae}} & flowering plant & 14 - 27 & Genetic & \citep{Wright1943}\\ 
  \makecell[l]{\textit{Borrichia}\\\textit{frutescens}} & salt marsh plant & 20 - 30 & Genetic+Survey & \citep{Antlfinger1982} \\ 
  \makecell[l]{\textit{Oreamnos}\\\textit{americanus}} & mountain goat & 36 - 100 & Genetic & \citep{Shirk2014} \\ 
  \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{Gainj- and Kalam- \\speaking people,\\Papua New Guinea} & 40 - 213 & Genetic & \citep{Rousset1997}\\ 
  \makecell[l]{\textit{Formica sp.}} & colonial ants & 50 - 100 & Genetic & \citep{Pamilo1983} \\ 
  \makecell[l]{\textit{Astrocaryum}\\\textit{mexicanum}} & palm tree & 102 - 895 & Genetic+survey & \citep{Eguiarte1993} \\ 
  \makecell[l]{\textit{Spermophilus}\\\textit{mollis}} & ground squirrel & 204 - 480 & Genetic+Survey & \citep{Antolin2001}\\
  \makecell[l]{\textit{Sceloperus}\\\textit{olivaceus}} & lizard & 225 - 270 & Survey & \citep{Kerster1964} \\ 
  \makecell[l]{\textit{Dieffenbachia}\\\textit{longispatha}} & \makecell[l]{beetle-pollinated\\colonial herb} & 227 - 611 & Survey & \citep{Young1988} \\ 
  \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{Gainj- and Kalam- \\speaking people,\\Papua New Guinea} & 410 & Survey & \citep{Rousset1997} \\ 
  \makecell[l]{\textit{Quercus}\\\textit{laevis}} & Oak tree & $>440$ & Genetic & \citep{Berg1995} \\ 
  \makecell[l]{\textit{Drosophila}\\\textit{pseudoobscura}} & fruit fly & 500 - 1,000 & Survey+Crosses & \citep{Wright1946} \\ 
  \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{POPRES data\\NE Europe} & 1,342 - 5,425 & Genetic & \citep{Ringbauer2017} \\ 
  \makecell[l]{\textit{Bebicium}\\\textit{vittatum}} & intertidal snail & 240,000 & Survey & \citep{Rousset1997}\\ 
  \makecell[l]{\textit{Bebicium}\\\textit{vittatum}} & intertidal snail & 360,000 & Genetic & \citep{Rousset1997}\\ 
   \hline
\end{tabular}
\label{table:NStable}
\end{table}
